# Script to try out cilia beat frequency analysis in R
# Sam Siljee
# 19 February 2025

---
title: "CBF analysis"
author: "Sam Siljee"
date: '2022-06-17'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

In this script I'm using the variable `d` to represent the donor/well, `x` for the pixel x coordinate, `y` for the pixel y coordinate, and `z` for the frame number.

```{r packages, include = FALSE}
library(dplyr) # Pipe and data manipulation
library(ggplot2) # plotting
library(signal) # Windowing for FFT
library(tiff) # Read in .tiff files
library(png) # Write png images
library(stringr) # Strip ".MOV" from file names
library(tidyr) # For pivot_wider function in FFT data summary
library(EBImage) # For label function to identify clusters
```

```{r Set some constants}
# Directories
input_dir <- "Test_input/"
output_dir <- "Test_output/"

# Options for running the script
load_from_rda <- TRUE
write_tables <- FALSE
run_signal_power <- FALSE
run_variance <- FALSE
run_sd <- FALSE
run_phase_variance <- FALSE
run_shannon_phase_entropy <- FALSE
run_MLAD <- FALSE
run_cluster_phase <- FALSE

# Set video details
height <- 360
width <- 640
frame_rate <- 480
num_frames <- 1024

# Minimum frequency threshold (Hz)
frequency_threshold <- 5
lower_threshold <- 7.7
upper_threshold <- 15.5
cluster_size_threshold <- 10

# LUT file
annotations_file <- "annotations.tsv"
```

```{r Setup}
# Vector for donor IDs - read from available .MOV files in input folder
donor_IDs <- list.files(input_dir) %>%
  grep(pattern = ".MOV$", value = TRUE) %>%
  str_remove(".MOV")

# Read in annotations
annotations <- read.delim(paste0(input_dir, annotations_file))

# Load functions
source("Functions.R")
```

## Run cilia analysis
Changed to run the whole analysis on a donor by donor basis to speed up processing. only saving the 3D arrays from the video input rather than saving data from every intermittent step.
```{r Run cilia beat frequency analysis - whole process by donor}
# Initialise list to save unconverted FFT data for all donors
FFT_list_unconverted <- list()

# Initialise list to save converted FFT data for all donors
FFT_list_converted <- list()

# Repeat process for each donor
for (d in donor_IDs) {
  # Print message to keep track of progress
  print(
    paste0(
      "Starting on donor ",
      d,
      "; ",
      which(donor_IDs %in% d),
      " of ",
      length(donor_IDs)
    )
  )
  
  ## Extract frames / load data ----
  if (load_from_rda) {
    # Load from prepared data
    print("Loading .rda")
    load(paste0(output_dir, "prepared_data/", d, "_3D_array.rda"))
  } else {
    # Read in from video data
    print("Extracting frames")
    
    # Use ffmpeg to extract all frames as single-channel greyscale TIFF (lossless)
    cmd <- paste0(
      "ffmpeg -i ",
      input_dir,
      d,
      ".MOV -pix_fmt gray -compression_algo lzw -vsync 0 ",
      output_dir,
      "temp_frames/frame_%d.tiff"
    )
    system(cmd)
    
    # Initialize an empty 3D array
    video_array <- array(NA, dim = c(height, width, num_frames))
    
    # Read in .png frames and select one channel (gray scale)
    for (z in 1:num_frames) {
      # Read in the frame
      video_array[, , z] <- readTIFF(paste0(output_dir, "temp_frames/frame_", z, ".tiff"), as.is = TRUE)
    }
    
    # Clear free space
    gc()
    
    # Remove temp frames from system
    cmd <- paste0(
      "rm ",
      output_dir,
      "temp_frames/frame_*.tiff"
    )
    system(cmd)
    
    # Save the data as a .rda file - for faster loading in future
    save(video_array, file = paste0(output_dir, "prepared_data/", d, "_3D_array.rda"))
  } # Close data loading else statement
  
  # Compensate for fluctuating background brightness
  frame_means <- numeric()
  for (z in 1:num_frames) {
    frame_means <- c(frame_means, mean(video_array[, , z]))
  }
  
  # Centre compensation around zero
  frame_means <- frame_means - mean(frame_means)
  
  # Subtract compensation from video array
  for (z in 1:num_frames) {
    video_array[, , z] <- video_array[, , z] - frame_means[z]
  }
  
  ## Signal power image ----
  if (run_signal_power) {
    print("Making signal power images")
    
    # Initialise blank matrix for signal power image
    signal_power_image <- matrix(data = 0, ncol = width, nrow = height)
    
    # Loop through frames to get differences
    for (z in 2:num_frames) {
      # Get absolute difference with previous frame
      difference <- abs(video_array[, , z] - video_array[, , (z - 1)])
      
      # Add to signal power image
      signal_power_image <- signal_power_image + difference
    }
    
    # Save as a PNG image - grayscale
    writePNG(normalise_image(signal_power_image), paste0(output_dir, "signal_power_images/signal_power_image_", d, ".png"))
    
    # Save as .tsv file
    if(write_tables) {
      write.table(
        signal_power_image,
        paste0(output_dir, "tsv_results/", d, "_signal_power_image.tsv"),
        row.names = FALSE,
        col.names = FALSE,
        sep = "\t"
      )
    }
    
    # Clear free space
    gc()
  }
  
  ## Variance image ----
  if (run_variance) {
    print("Making variance images")
    
    # Initialise blank matrix
    variance_matrix <- matrix(ncol = width, nrow = height)
    
    # Get variance by pixel
    for (y in 1:width) { # loop through columns
      for (x in 1:height) { # Loop through rows
        # Calculate variances and add to list in the required position
        variance_matrix[x, y] <- var(video_array[x, y, ])
      }
    }
    
    # Save as a PNG image - grayscale
    writePNG(normalise_image(variance_matrix), paste0(output_dir, "variance_images/variance_image_", d, ".png"))
    
    # Save as .tsv file
    if(write_tables) {
      write.table(
        variance_matrix,
        paste0(output_dir, "tsv_results/", d, "_variance_image.tsv"),
        row.names = FALSE,
        col.names = FALSE,
        sep = "\t"
      ) 
    }
  }
  
  ## SD ----
  if (run_sd) {
    print("Making SD images")
    
    # Initialise blank matrix
    sd_matrix <- matrix(ncol = width, nrow = height)
    
    # Get variance by pixel
    for (y in 1:width) { # loop through columns
      for (x in 1:height) { # Loop through rows
        # Calculate variances and add to list in the required position
        sd_matrix[x, y] <- sd(video_array[x, y, ])
      }
    }
    
    # Save as a PNG image - grayscale
    writePNG(normalise_image(sd_matrix), paste0(output_dir, "sd_images/sd_image_", d, ".png"))
    
    # Save as .tsv file
    if(write_tables) {
      write.table(
        sd_matrix,
        paste0(output_dir, "tsv_results/", d, "_sd_image.tsv"),
        row.names = FALSE,
        col.names = FALSE,
        sep = "\t"
      )
    }
  }
  
  ## FFT ----
  print("Making FFT images")
  
  # Create frequency vector
  freq <- seq(0, frame_rate / 2, length.out = num_frames / 2 + 1)
  
  # Calculate offset to exclude results below the threshold
  threshold_offset <- sum(freq < frequency_threshold)
  
  # Initialise matrices to contain dominant frequencies
  FFT_matrix <- matrix(ncol = width, nrow = height)
  FFT_matrix_unconverted <- FFT_matrix
  
  # Calculate FFT and extract dominant frequencies
  for (y in 1:width) { # loop through columns
    for (x in 1:height) { # Loop through rows
      # Get vector of pixel values
      pixel_values <- video_array[x, y, ]
      
      # Run FFT
      fft_result <- fft(pixel_values)
      
      # Get magnitudes - using only first half of FFT results
      magnitude <- abs(fft_result)[1:(num_frames / 2 + 1)]
      
      # Set all magnitudes below frequency threshold to -1 so they won't be selected
      magnitude[1:threshold_offset] <- -1
      
      # Find the frequency with maximum amplitude
      max_index <- which.max(magnitude)
      
      # Add to list in the required position - converted to real frequency and unconverted
      FFT_matrix[x, y] <- freq[max_index]
      FFT_matrix_unconverted[x, y] <- max_index
    }
  }
  
  # Add to list of all donors
  FFT_list_unconverted[[d]] <- FFT_matrix_unconverted
  FFT_list_converted[[d]] <- FFT_matrix
  
  # Save as a PNG image - grayscale
  writePNG(normalise_image(FFT_matrix), paste0(output_dir, "fft_images/FFT_freq_image_", d, ".png"))
  
  # Save as .tsv file
  if(write_tables) {
    write.table(
      FFT_matrix,
      paste0(output_dir, "tsv_results/", d, "_fft_image.tsv"),
      row.names = FALSE,
      col.names = FALSE,
      sep = "\t"
    ) 
  }
  
  # Save as a PNG image - grayscale
  writePNG(normalise_image(FFT_matrix_unconverted), paste0(output_dir, "fft_images/FFT_unconverted_freq_image_", d, ".png"))
  
  # Save as .tsv file
  if(write_tables) {
    write.table(
      FFT_matrix_unconverted,
      paste0(output_dir, "tsv_results/", d, "_fft_unconverted_image.tsv"),
      row.names = FALSE,
      col.names = FALSE,
      sep = "\t"
    ) 
  }
  
  # Clear free space
  gc()
  
  ## Identify clusters ----
  print("Identifying clusters")
  # Use custom function to identify cluster sizes
  cluster_size_matrix <- identify_clusters(FFT_matrix_unconverted)
  
  # Save as .tsv file
  write.table(
    cluster_size_matrix,
    paste0(output_dir, "tsv_results/", d, "_cluster_size_matrix.tsv"),
    row.names = FALSE,
    col.names = FALSE,
    sep = "\t"
  )
  
  # Save as a PNG image - grayscale
  writePNG(normalise_image(cluster_size_matrix), paste0(output_dir, "cluster_images/Cluster_size_image_", d, ".png"))
  
  # Create binary mask
  cluster_mask_unfilled <- cluster_size_matrix > cluster_size_threshold
  
  # Save as .tsv file
  write.table(
    cluster_mask_unfilled,
    paste0(output_dir, "tsv_results/", d, "_cluster_mask_unfilled.tsv"),
    row.names = FALSE,
    col.names = FALSE,
    sep = "\t"
  )
  
  # Save as a PNG image - grayscale
  writePNG(normalise_image(cluster_mask_unfilled), paste0(output_dir, "cluster_images/Cluster_mask_unfilled_", d, ".png"))
  
  # Fill the cluster mask
  cluster_mask <- fill_clusters(cluster_size_matrix) > cluster_size_threshold
  
  # Save as .tsv file
  write.table(
    cluster_mask,
    paste0(output_dir, "tsv_results/", d, "_cluster_mask.tsv"),
    row.names = FALSE,
    col.names = FALSE,
    sep = "\t"
  )
  
  # Save as a PNG image - grayscale
  writePNG(normalise_image(cluster_mask), paste0(output_dir, "cluster_images/Cluster_mask_", d, ".png"))
  
  ## Histograms ----
  print("Making histograms")
  
  # Histogram of converted frequencies - filtered by cluster mask
  ggplot(data.frame(Frequency = as.numeric(FFT_matrix[cluster_mask])), aes(x = Frequency)) +
    geom_histogram(binwidth = 1) +
    labs(
      x = "Frequency (Hz)",
      y = "Count",
      title = paste("Hisotgram of dominant frequency:", d)
    ) +
    theme_bw()
  
  # Save the plot
  ggsave(paste0(output_dir, "histograms/", d, " frequency histogram.png"), width = 8, height = 5.5)
  
  # Histogram of unconverted (indexes)
  ggplot(data.frame(Index = as.numeric(FFT_matrix_unconverted[cluster_mask])), aes(x = Index)) +
    geom_histogram(binwidth = 1) +
    labs(
      x = "Dominant frequency index (binwidth = 1)",
      y = "Count",
      title = paste("Hisotgram of dominant frequency indexes:", d)
    ) +
    theme_bw()
  
  # Save the plot
  ggsave(paste0(output_dir, "histograms/", d, " frequency index histogram.png"), width = 8, height = 5.5)
  
  ## Phase variance ----
  if(run_phase_variance) {
    
  }
  
  # Shannon phase entropy ----
  if(run_shannon_phase_entropy) {
    
  }
  
  # MLAD, Mean Local Angular Deviation ----
  if(run_MLAD) {
    
  }
  
  # Fraction of cluster pixels that have at least one neighbor with a phase difference of at least "some threshold" ----
  if(run_cluster_phase) {
    
  }
  
} # Close loop over donors

# Save the FFT data as .rda files
save(FFT_list_unconverted, file = paste0(output_dir, "FFT_list_unconverted.rda"))
save(FFT_list_converted, file = paste0(output_dir, "FFT_list_converted.rda"))
```
