# Script to try out cilia beat frequency analysis in R
# Sam Siljee
# 19 February 2025

---
  title: "CBF analysis"
author: "Sam Siljee"
date: '2022-06-17'
output: html_document
---
  
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

In this script I'm using the variable `d` to represent the donor/well, `x` for the pixel x coordinate, `y` for the pixel y coordinate, and `f` for the frame number.

```{r packages, include = FALSE}
library(dplyr) # Pipe and data manipulation
library(ggplot2) # plotting
library(signal) # Windowing for FFT
library(magick) # Read in video files, with `ffmpeg`
library(stringr) # Strip ".MOV" from file names
```

```{r Set some constants}
# Directories
input_dir <- "Input/"
output_dir <- "Output/"

# Set video details
height <- 360
width <- 640
frame_rate <- 480
num_frames <- 512

# Vector for donor IDs - read from available .MOV files in input
donor_IDs <- list.files(input_dir) %>%
  grep(pattern = ".MOV$", value = TRUE) %>%
  str_remove(".MOV")
```

## Extract frames
```{r Extract frames}
# Initialise a list to hold all pixel data
data_list <- list()

# Iterate over donors - extract relevant pixel values
for (d in donor_IDs) {
  # Print message to keep track of progress
  print(
    paste0(
      "Extracting frames for donor ",
      d,
      ", ",
      which(donor_IDs %in% d),
      " of ",
      length(donor_IDs)
    )
  )

  # Load video with magick
  video_frames <- image_read_video(paste0(input_dir, d, ".MOV"), fps = NULL)

  # keep the last X frames
  video_frames <- video_frames[(length(video_frames) - (num_frames - 1)):length(video_frames)]

  # Initialize an empty list to store grayscale frames
  frames_array <- array(NA, dim = c(width, height, num_frames))

  # Extract pixel values and convert to greyscale
  for (f in 1:num_frames) {
    # Extract the current frame
    frame <- video_frames[f]

    # Convert the frame to grayscale
    frame_gray <- image_convert(frame, colorspace = "gray")

    # Get pixel values and scale to 0-255 range
    frame_matrix <- as.integer(image_data(frame_gray))

    # Store the frame as a matrix in the list
    frames_array[,,f] <- matrix(frame_matrix, ncol = width)
  }

  # Save the frame list under the relevant donor
  data_list[[d]] <- frames_array

  # Clear free space
  rm(frames_array)
  gc()
}

# Save the data as a .rda file
save(data_list, file = paste0(output_dir, "data_list.rda"))
```

## Signal power image
```{r Signal power image}
# A kind of signal power image:
# Start loop through all donors
for (d in donor_IDs) {
  # Progress monitor
  print(
    paste0(
      "Maing signal power images for donor ",
      d,
      ", ",
      which(donor_IDs %in% d),
      " of ",
      length(donor_IDs)
    )
  )

  # Initialise blank matrix for signal power image
  signal_power_image <- matrix(data = 0, ncol = width, nrow = height)

  # Loop through frames to get differences
  for (f in 2:num_frames) {
    # Get absolute difference with previous frame
    difference <- abs(data_list[[d]][,,f] - data_list[[d]][,,(f-1)])
    # Add to signal power image
    signal_power_image <- signal_power_image + difference
  }
  
  # Scale manually to 0-255
  max_signal <- max(signal_power_image)
  signal_power_image <- round(signal_power_image / max_signal * 255)

  # Save as a PNG image - grayscale
  png(paste0(output_dir, "signal_power_images/signal_power_image_", d, "_grayscale.png"), width = width, height = height)

  image(signal_power_image, col = gray.colors(256), axes = FALSE)

  dev.off()

  # Save as a PNG image - rainbow
  png(paste0(output_dir, "signal_power_images/signal_power_image_", d, "_rainbow.png"), width = width, height = height)

  image(signal_power_image, col = rainbow(256), axes = FALSE)

  dev.off()
} # End loop through donors
```

## Variances
```{r Variances}
# Initialise list to contain variances
variance_list <- vector("list", length(donor_IDs))
names(variance_list) <- donor_IDs
for (d in donor_IDs) {
  variance_list[[d]] <- matrix(ncol = width, nrow = height)
}

for (d in donor_IDs) { # Loop through donors
  # Progress monitor
  print(
    paste0(
      "Calculating pixel variances for donor ",
      d,
      ", ",
      which(donor_IDs %in% d),
      " of ",
      length(donor_IDs)
    )
  )

  for (x in 1:height) { # loop through rows
    for (y in 1:width) { # Loop through columns
      # Calculate variances and add to list in the required position
      variance_list[[d]][x, y] <- var(data_list[[d]][x, y, ])
    }
  }
}


# Save the data as a .rda file
save(variance_list, file = paste0(output_dir, "variance_list.rda"))

# Create variance images
for (d in donor_IDs) {
  # Progress monitor
  print(
    paste0(
      "Creating pixel variance images for donor ",
      d,
      ", ",
      which(donor_IDs %in% d),
      " of ",
      length(donor_IDs)
    )
  )

  # Save as a PNG image - grayscale
  png(paste0(output_dir, "variance_images/variance_image_", d, "_grayscale.png"), width = width, height = height)

  image(variance_list[[d]], col = gray.colors(256), axes = FALSE, zlim = range(variance_list[[d]]))

  dev.off()

  # Save as a PNG image - rainbow
  png(paste0(output_dir, "variance_images/variance_image_", d, "_rainbow.png"), width = width, height = height)

  image(variance_list[[d]], col = rainbow(256), axes = FALSE, zlim = range(variance_list[[d]]))

  dev.off()
}
```

## FFT dominant frequency
```{r FFT frequencies}
# Create frequency vector
freq <- seq(0, frame_rate / 2, length.out = num_frames / 2 + 1)

# Initialise lists to contain dominant frequencies
FFT_freq_list <- vector("list", length(donor_IDs))
names(FFT_freq_list) <- donor_IDs
for (d in donor_IDs) {
  FFT_freq_list[[d]] <- matrix(ncol = width, nrow = height)
}
FFT_freq_list_unconverted <- FFT_freq_list

for (d in donor_IDs) { # Loop through donors
  # Progress monitor
  print(
    paste0(
      "Calculating FFT maximum frequency for donor ",
      d,
      "; ",
      which(donor_IDs %in% d),
      " of ",
      length(donor_IDs)
    )
  )

  for (x in 1:height) { # loop through rows
    for (y in 1:width) { # Loop through columns
      # Get vector of pixel values
      pixel_values <- data_list[[d]][x, y, ]

      # Calculate the dominant frequency from pixel values
      # Run FFT
      fft_result <- fft(pixel_values)

      # Find maximum magnitude - using half the dataset (symetrical)
      magnitude <- abs(fft_result[1:(num_frames / 2 + 1)])

      # Find the frequency with maximum amplitude
      # Excluding the DC component (first element)
      max_index <- which.max(magnitude[-1]) + 1

      # Add to list in the required position - converted to real frequency
      FFT_freq_list[[d]][x, y] <- freq[max_index]
      
      # Add to list in the required position - unconverted to real frequency
      FFT_freq_list_unconverted[[d]][x, y] <- max_index
    }
  }
}

# Save the data as .rda files
save(FFT_freq_list, file = paste0(output_dir, "FFT_freq_list.rda"))

save(FFT_freq_list_unconverted, file = paste0(output_dir, "FFT_freq_list_unconverted.rda"))

# Create dominant frequency images
for (d in donor_IDs) {
  # Progress monitor
  print(
    paste0(
      "Creating pixel FFT frequency image for donor ",
      d,
      ", ",
      which(donor_IDs %in% d),
      " of ",
      length(donor_IDs)
    )
  )

  # Save as a PNG image - grayscale
  png(paste0(output_dir, "fft_images/FFT_freq_image_", d, "_grayscale.png"), width = width, height = height)

  image(FFT_freq_list[[d]], col = gray.colors(256), axes = FALSE, zlim = range(FFT_freq_list[[d]]))

  dev.off()

  # Save as a PNG image - rainbow
  png(paste0(output_dir, "fft_images/FFT_freq_image_", d, "_rainbow.png"), width = width, height = height)

  image(FFT_freq_list[[d]], col = rainbow(256), axes = FALSE, zlim = range(FFT_freq_list[[d]]))

  dev.off()
  
  # Save as a PNG image - grayscale
  png(paste0(output_dir, "fft_images/FFT_unconverted_freq_image_", d, "_grayscale.png"), width = width, height = height)

  image(FFT_freq_list_unconverted[[d]], col = gray.colors(256), axes = FALSE, zlim = range(FFT_freq_list_unconverted[[d]]))

  dev.off()

  # Save as a PNG image - rainbow
  png(paste0(output_dir, "fft_images/FFT_unconverted_freq_image_", d, "_rainbow.png"), width = width, height = height)

  image(FFT_freq_list_unconverted[[d]], col = rainbow(256), axes = FALSE, zlim = range(FFT_freq_list_unconverted[[d]]))

  dev.off()
}
```

## Histograms of indexes
```{r histograms}
# Create dominant index histograms
for (d in donor_IDs) {
  # Progress monitor
  print(
    paste0(
      "Creating dominant frequency histogram for donor ",
      d,
      ", ",
      which(donor_IDs %in% d),
      " of ",
      length(donor_IDs)
    )
  )

  indexes <- FFT_freq_list_unconverted[[d]] %>% as.numeric()

  ggplot(data.frame(Index = indexes), aes(x = Index)) +
    geom_histogram(binwidth = 1) +
    labs(
      x = "Dominant frequency index (binwidth = 1)",
      y = "Count",
      title = paste("Hisotgram of dominant frequency indexes:", d)
    ) +
    theme_bw()

  # Save the plot
  ggsave(paste0(output_dir, "histograms/", d, " frequency index histogram.png"), width = 8, height = 5.5)
}
```
