# Script to try out cilia beat frequency analysis in R
# Sam Siljee
# 19 February 2025

---
title: "CBF analysis"
author: "Sam Siljee"
date: '2022-06-17'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

In this script I'm using the variable `d` to represent the donor/well, `x` for the pixel x coordinate, `y` for the pixel y coordinate, and `z` for the frame number.

```{r packages, include = FALSE}
library(dplyr) # Pipe and data manipulation
library(ggplot2) # plotting
library(signal) # Windowing for FFT
library(png) # Read in .png files
library(stringr) # Strip ".MOV" from file names
```

```{r Set some constants}
# Directories
input_dir <- "Test_input/"
output_dir <- "Test_output/"

# Load data from pre-prepared data
load_from_rda <- TRUE

# Set video details
height <- 360
width <- 640
frame_rate <- 480
num_frames <- 512
last_frame <- 9000

# Minimum frequency threshold (Hz)
frequency_threshold <- 5

# Vector for donor IDs - read from available .MOV files in input folder
donor_IDs <- list.files(input_dir) %>%
  grep(pattern = ".MOV$", value = TRUE) %>%
  str_remove(".MOV")
```

## Run cilia analysis
Changed to run the whole analysis on a donor by donor basis to speed up processing. only saving the 3D arrays from the video input rather than saving data from every intermittent step.
```{r Run cilia beat frequency analysis - whole process by donor}
# Repeat process for each donor
for (d in donor_IDs) {
  # Print message to keep track of progress
  print(
    paste0(
      "Starting on donor ",
      d,
      ", ",
      which(donor_IDs %in% d),
      " of ",
      length(donor_IDs)
    )
  )
  
  # Time for optimisation
  start_time <- Sys.time()
  
  ## Extract frames / load data ----
  if(load_from_rda) {
    # Load from prepared data
    print("Loading .rda")
    load(paste0(output_dir, "prepared_data/", d, "_3D_array.rda"))
    
  } else {
    # Read in from video data
    print("Extracting frames")
    
    # Use ffmpeg to extract frames and write to .png
    cmd <- paste0(
      "ffmpeg -i ",
      input_dir,
      d,
      ".MOV -vf \"select=between(n\\,",
      last_frame - num_frames + 1,
      "\\,",
      last_frame,
      ")\" -vsync 0 ",
      output_dir,
      "temp_frames/frame_%d.png"
    )
    system(cmd)
    
    # Initialize an empty 3D array
    video_array <- array(NA, dim = c(height, width, num_frames))
    
    # Read in .png frames and select one channel (gray scale)
    for (z in 1:num_frames) {
      #Read in the frame
      video_array[,,z] <- readPNG(paste0(output_dir, "temp_frames/frame_", z, ".png"))[,,1]
    }
    
    # Clear free space
    gc()
    
    # Save the data as a .rda file - for faster loading in future
    save(video_array, file = paste0(output_dir, "prepared_data/", d, "_3D_array.rda"))
  } # Close data loading else statement
  
  # Video load time
  video_load_time <- Sys.time() - start_time
  print(paste(
    "Video loading time:",
    video_load_time
  ))
  
  ## Signal power image ----
  print("Making signal power images")
  
  # Initialise blank matrix for signal power image
  signal_power_image <- matrix(data = 0, ncol = width, nrow = height)
  
  # Loop through frames to get differences
  for (z in 2:num_frames) {
    # Get absolute difference with previous frame
    difference <- abs(video_array[,,z] - video_array[,,(z-1)])
    
    # Add to signal power image
    signal_power_image <- signal_power_image + difference
  }
  
  # Scale manually to 0-255
  signal_power_image <- round(signal_power_image / max(signal_power_image) * 255)
  
  # Save as a PNG image - grayscale
  png(paste0(output_dir, "signal_power_images/signal_power_image_", d, "_grayscale.png"), width = width * 10, height = height * 10)
  par(mar = c(0, 0, 0, 0))
  image(signal_power_image, col = gray.colors(256), axes = FALSE)
  dev.off()
  
  # Save as a PNG image - rainbow
  png(paste0(output_dir, "signal_power_images/signal_power_image_", d, "_rainbow.png"), width = width * 10, height = height * 10)
  par(mar = c(0, 0, 0, 0))
  image(signal_power_image, col = rainbow(256), axes = FALSE)
  dev.off()
  
  # Save as .tsv file
  write.table(
    matrix(signal_power_image, nrow = height, ncol = width),
    paste0(output_dir, "tsv_results/", d, "_signal_power_image.tsv"),
    row.names = FALSE,
    col.names = FALSE,
    sep = "\t")
  
  # Clear free space
  gc()
  
  ## Variance image ----
  print("Making variance images")
  
  # Initialise blank matrix
  variance_matrix <- matrix(ncol = width, nrow = height)
  
  # Get variance by pixel
  for (y in 1:width) { # loop through columns
    for (x in 1:height) { # Loop through rows
      # Calculate variances and add to list in the required position
      variance_matrix[x, y] <- var(video_array[x, y, ])
    }
  }
  
  # Scale manually to 0-255
  variance_image <- round(variance_matrix / max(variance_matrix) * 255)
  
  # Save as a PNG image - grayscale
  png(paste0(output_dir, "variance_images/variance_image_", d, "_grayscale.png"), width = width * 10, height = height * 10)
  par(mar = c(0, 0, 0, 0))
  image(variance_image, col = gray.colors(256), axes = FALSE)
  dev.off()
  
  # Save as a PNG image - rainbow
  png(paste0(output_dir, "variance_images/variance_image_", d, "_rainbow.png"), width = width * 10, height = height * 10)
  par(mar = c(0, 0, 0, 0))
  image(variance_image, col = rainbow(256), axes = FALSE)
  dev.off()
  
  # Save as .tsv file
  write.table(
    matrix(variance_image, nrow = height, ncol = width),
    paste0(output_dir, "tsv_results/", d, "_variance_image.tsv"),
    row.names = FALSE,
    col.names = FALSE,
    sep = "\t")
  
  ## FFT ----
  print("Making FFT images")
  
  # Create frequency vector
  freq <- seq(0, frame_rate / 2, length.out = num_frames / 2 + 1)
  
  # Calculate offset to exclude results below the threshold
  threshold_offset <- sum(freq < frequency_threshold)
  
  # Initialise matrices to contain dominant frequencies
  FFT_matrix <- matrix(ncol = width, nrow = height)
  FFT_matrix_unconverted <- FFT_matrix
  
  # Calculate FFT and extract dominant frequencies
  for (y in 1:width) { # loop through columns
    for (x in 1:height) { # Loop through rows
      # Get vector of pixel values
      pixel_values <- video_array[x, y, ]
      
      # Run FFT
      fft_result <- fft(pixel_values)
      
      # Get magnitudes - using only first half of FFT results
      magnitude <- abs(fft_result)[1:(num_frames/2 + 1)]
      
      # Set all magnitudes below frequency threshold to -1 so they won't be selected
       magnitude[1:threshold_offset] <- -1
      
      # Find the frequency with maximum amplitude
      max_index <- which.max(magnitude)
      
      # Add to list in the required position - converted to real frequency and unconverted
      FFT_matrix[x, y] <- freq[max_index]
      FFT_matrix_unconverted[x, y] <- max_index
    }
  }
  
  # Scale manually to 0-255
  fft_image <- round(FFT_matrix / max(FFT_matrix) * 255)
  fft_unconverted_image <- round(FFT_matrix_unconverted / max(FFT_matrix_unconverted) * 255)
  
  # Save as a PNG image - grayscale
  png(paste0(output_dir, "fft_images/FFT_freq_image_", d, "_grayscale.png"), width = width * 10, height = height * 10)
  par(mar = c(0, 0, 0, 0))
  image(fft_image, col = gray.colors(256), axes = FALSE)
  dev.off()
  
  # Save as a PNG image - rainbow
  png(paste0(output_dir, "fft_images/FFT_freq_image_", d, "_rainbow.png"), width = width * 10, height = height * 10)
  par(mar = c(0, 0, 0, 0))
  image(fft_image, col = rainbow(256), axes = FALSE)
  dev.off()
  
  # Save as .tsv file
  write.table(
    matrix(fft_image, nrow = height, ncol = width),
    paste0(output_dir, "tsv_results/", d, "_fft_image.tsv"),
    row.names = FALSE,
    col.names = FALSE,
    sep = "\t")
  
  # Save as a PNG image - grayscale
  png(paste0(output_dir, "fft_images/FFT_unconverted_freq_image_", d, "_grayscale.png"), width = width * 10, height = height * 10)
  par(mar = c(0, 0, 0, 0))
  image(fft_unconverted_image, col = gray.colors(256), axes = FALSE)
  dev.off()
  
  # Save as a PNG image - rainbow
  png(paste0(output_dir, "fft_images/FFT_unconverted_freq_image_", d, "_rainbow.png"), width = width * 10, height = height * 10)
  par(mar = c(0, 0, 0, 0))
  image(fft_unconverted_image, col = rainbow(256), axes = FALSE)
  dev.off()
  
  # Save as .tsv file
  write.table(
    matrix(fft_unconverted_image, nrow = height, ncol = width),
    paste0(output_dir, "tsv_results/", d, "_fft_unconverted_image.tsv"),
    row.names = FALSE,
    col.names = FALSE,
    sep = "\t")
  
  # Clear free space
  gc()
  
  ## Histograms ----
  print("Making histograms")
  
  # Histogram of converted frequencies
  ggplot(data.frame(Frequency = as.numeric(FFT_matrix)), aes(x = Frequency)) +
    geom_histogram(binwidth = 1) +
    labs(
      x = "Frequency (Hz)",
      y = "Count",
      title = paste("Hisotgram of dominant frequency:", d)
    ) +
    theme_bw()
  
  # Save the plot
  ggsave(paste0(output_dir, "histograms/", d, " frequency histogram.png"), width = 8, height = 5.5)
  
  # Histogram of unconverted (indexes)
  ggplot(data.frame(Index = as.numeric(FFT_matrix)), aes(x = Index)) +
    geom_histogram(binwidth = 1) +
    labs(
      x = "Dominant frequency index (binwidth = 1)",
      y = "Count",
      title = paste("Hisotgram of dominant frequency indexes:", d)
    ) +
    theme_bw()
  
  # Save the plot
  ggsave(paste0(output_dir, "histograms/", d, " frequency index histogram.png"), width = 8, height = 5.5)
  
  # Video processing time
  video_processing_time <- Sys.time() - start_time - video_load_time
  print(paste(
    "Video processing time:",
    video_processing_time
  ))
} # Close loop over donors
```
