---
title: "CBF_ciliaR"
author: "Sam Siljee"
date: "2025-04-06"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Setup
Script to analyse cilia beat frequency using the ciliaR package from Oriane Grant et. al. (Dr Claire Smith lab), paper available:
https://doi.org/10.1101/2023.12.20.572306
```{r libraries}
library(dplyr) # Data manipulation
library(stringr) # Get list of donors from input directory
library(ggplot2) # Plotting
library(magick) # Read in video files, with `ffmpeg`
source("analyse_cilia2 (PC).R") # Get the analyse_cilia2 function
```

```{r Set some variables}
# Directories
input_dir <- "Input/"
output_dir <- "Output/"

# Read in data fresh
read_from_rda <- TRUE

# Set video details
height <- 360
width <- 640
frame_rate <- 480
num_frames <- 2^9 # Use powers of 2 for analysis

# Vector for donor IDs - read from available .MOV files in input
donor_IDs <- list.files(input_dir) %>%
  grep(pattern = ".MOV$", value = TRUE) %>%
  str_remove(".MOV")
```

## Load the data
Read in from .mov files into a list of data
```{r Extract frames}
if (read_from_rda) {
  load(paste0(output_dir, "data_list.rda"))
} else {
  # Initialise a list to hold all pixel data
  data_list <- list()

  # Iterate over donors - extract relevant pixel values
  for (d in donor_IDs) {
    # Print message to keep track of progress
    print(
      paste0(
        "Extracting frames for donor ",
        d,
        ", ",
        which(donor_IDs %in% d),
        " of ",
        length(donor_IDs)
      )
    )

    # Load video with magick
    video_frames <- image_read_video(paste0(input_dir, d, ".MOV"), fps = NULL)

    # keep the last X frames
    video_frames <- video_frames[(length(video_frames) - (num_frames - 1)):length(video_frames)]

    # Initialize an empty list to store grayscale frames
    frames_list <- vector("list", num_frames)

    # Extract pixel values and convert to greyscale
    for (f in 1:num_frames) {
      # Extract the current frame
      frame <- video_frames[f]

      # Convert the frame to grayscale
      frame_gray <- image_convert(frame, colorspace = "gray")

      # Get pixel values and scale to 0-255 range
      frame_matrix <- as.integer(image_data(frame_gray))

      # Store the frame as a matrix in the list
      frames_list[[f]] <- matrix(frame_matrix, ncol = width)
    }

    # Save the frame list under the relevant donor
    data_list[[d]] <- frames_list

    # Clear the cache to free space
    gc()
  }

  # Save the data as a .rda file
  save(data_list, file = paste0(output_dir, "data_list.rda"))
}
```

## Format data
Format for `analyse_cilia2` function. The format takes the form of a column for every ROI. In the paper this takes the form of dividing intoa  set number of rows and columns, to a minimum of 1,600 ROIs. I use every pixel as its own ROI. every row represents the mean intensity of the ROIs, one row for every frame. The number of frames used must be a power of 2 (eg. 256, 512). This is then saved as a .txt file The first column is a set of row index numbers, the second column called "Label" is the file name, subsequent columns are labelled 
```{r Format the data for ciliaR}
for(d in donor_IDs) {
  # Progress monitor
  print(
    paste0(
      "Formatting data for donor ",
      d,
      ", ",
      which(donor_IDs %in% d),
      " of ",
      length(donor_IDs)
    )
  )
  
  # Subset the data by donor
  data <- data_list[[d]]
  
  # Initialise the data.frame to fill
  formatted_data <- matrix(NA, nrow = num_frames, ncol = height*width)
  
  # Add in the data
  for (f in 1:num_frames) {
    formatted_data[f,] <- as.numeric(data[[f]])
  }
  
  # Change to data.frame format to change column names and add Label column
  formatted_data <- as.data.frame(formatted_data)
  Label <- rep(d, num_frames)
  formatted_data <- cbind(Label, formatted_data)
  colnames(formatted_data) <- c("Label", paste0("Mean", 1:(height*width)))
  
  # Save as .txt file
  write.table(
    formatted_data,
    file = paste0(input_dir, "Formatted_for_ciliaR/", d, ".txt"),
    quote = FALSE,
    row.names=TRUE,
    sep="\t"
  )
}
```

## Run ciliaR
Run the ciliaR analysis - default noise and upper limit thresholds
```{r ciliaR}
# # Run using the code suggested, currently not functioning
# list.files(paste0(input_dir, "Formatted_for_ciliaR"))
# 
# analyse_cilia2(
#   path = paste0(input_dir, "Formatted_for_ciliaR"),
#   FRate = frame_rate,
#   NFrame = num_frames,
#   NoiseLevel = 3,
#   UpperLimit = 30
# )

# Run using the code taken from the ciliaR function code
# Set the variables
FRate <- frame_rate
NFrame <- num_frames
NoiseLevel <- 3
UpperLimit <- 30

# Code chunk  - ciliR analysis part 1
dataset <- data.frame()
datalist = list()

FreqRes <- FRate / NFrame
BackGr <- round(NoiseLevel/FreqRes, digits = 0)
BackGr1 <- BackGr + 1
EndFrame <- round(UpperLimit/FreqRes, digits = 0)

# Start loop through donors
for(d in "21_073_1") {
  # Progress monitor
  print(
    paste0(
      "Running ciliaR for donor ",
      d,
      ", ",
      which(donor_IDs %in% d),
      " of ",
      length(donor_IDs)
    )
  )
  
  # Load the data
  my_data <- utils::read.delim(paste0(input_dir, "Formatted_for_ciliaR/", d, ".txt"),  header = TRUE, stringsAsFactors = FALSE, check.names = FALSE)
  
  # # Trim number of ROIs for faster processing
  # my_data <- my_data[,1:1600]
  
  #fourier transform
  my_data2 <- head(my_data, NFrame)
  my_data3 <- my_data2[, -c(1:2), -c(100001)] # delete first columns which are not of interest
  fourierd <- apply(my_data3,2,function(x) fft(as.numeric(x)))
  fourierdat <- as.data.frame(fourierd)
  
  #get absolute values
  AbsAll <- abs(fourierdat)
  
  #noise reduction: compare the amplitude peak of the data to the amplitude peak of the background noise
  AbsNoise <- AbsAll[2:BackGr,]    # divide the data into subsets of background noise (<3Hz)
  AbsReal <- AbsAll[BackGr1:EndFrame,]
  max.power.F1<- apply(AbsNoise, 2, max) # find the maximum value of each subset
  max.power.F2<- apply(AbsReal, 2, max) # find the maximum value of each subset
  
  #if F2 >F1 find row number of F2
  BF <- sapply(AbsReal[-1, ], which.max) ## find the row number containing the max. value of each column
  BF2 <- as.numeric(BF)
  BFData <- ifelse(max.power.F2 > (max.power.F1 * NoiseLevel),  BF2 , 0)
  
  #convert to Hz
  sapply(BFData, class)
  BFdata2 <- as.numeric(BFData) * FreqRes # times row number by frequency resolution
  Hz <- BFdata2 + FreqRes # add value of missing row
  
  #only include data in clinical range
  BFData4 <- ifelse(Hz>3 & Hz<UpperLimit, Hz, 0)   ##
  dat <- data.frame(BFData4)
  names(dat)[1] <- d
  datalist[[d]] <- dat
}

#present big results table
bigdata <- suppressMessages(dplyr::bind_cols(datalist))
bigdata[bigdata==0] <- NA

# Save results as .csv
write.csv(bigdata, file = paste0(output_dir, "summary.csv"), row.names = FALSE)
print(summary(bigdata))
assign("ciliaR_results", bigdata, envir = .GlobalEnv)

# Save as .rda file
save(ciliaR_results, file = paste0(input_dir, "ciliaR_results.rda"))
```

## Write to .csv for processing in ImageJ
```{r Write to .csv}
for(d in donor_IDs) {
  write.csv(
    matrix(ciliaR_results[[d]], nrow = height, ncol = width),
    paste0(output_dir, "csv_results/", d, "_fft_dominant_frequencies.csv"))
}
```

